---
title: "Example Map-Matching Trajectories"
format: html
---

```{r}
library(tidyverse)
library(sf)
library(leaflet)
library(here)
library(glue)
```

# Example from Newson and Krumm (2009)

How to download the data:

1. Navigate to https://www.microsoft.com/en-us/research/publication/hidden-markov-map-matching-noise-sparseness/

2. Save the test trip GPS data text file to your machine as `hmm_test_trip_data.txt`.

3. Save the road network data text file to your machine as `hmm_road_network.txt`.

4. Save the ground truth path text file to your machine as `hmm_ground_truth.txt`.

This example dataset came from a drive in Seattle, Washington, USA.

```{r}
sea_trip <- read_tsv(here("hmm_test_trip_data.txt"), col_names = TRUE) |>
  select(`Date (UTC)`, `Time (UTC)`, Latitude, Longitude)
```

```{r}
sea_trip_sf <- sea_trip |>
  st_as_sf(coords = c("Longitude", "Latitude"), 
           crs = 4326)
```

```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data = sea_trip_sf, label = ~glue("{`Time (UTC)`}"))
```

```{r}
sea_roads <- read_delim(here("hmm_road_network.txt"), col_names = TRUE)
```

```{r}
sea_roads_geometry <- sea_roads$`LINESTRING()` |>
  map(\(x) st_as_sfc(x, crs = 4326))

sea_roads_geometry_listc <- list_c(sea_roads_geometry)

sea_roads_sf <- sea_roads |>
  select(!`LINESTRING()`) |>
  st_set_geometry(sea_roads_geometry_listc)
```

```{r}
sea_route <- read_delim(here("hmm_ground_truth.txt"), col_names = TRUE)

sea_route_sf <- left_join(sea_route, sea_roads_sf, by = "Edge ID") |>
  st_as_sf(crs = 4326)
```

```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data = sea_trip_sf, label = ~glue("{`Time (UTC)`}"), color = "blue") |>
  addPolylines(data = sea_route_sf, label = ~glue("{`Edge ID`}"), color = "gray")
```


# Examples from Kubička et al (2015)

How to download the data:

1. Navigate to https://zenodo.org/records/57731

2. Download `map-matching-dataset.zip` to your machine.

3. Unzip `map-matching-dataset.zip`.

The 100 example datasets came from a public collection called Planet GPX, which is part of the Open Street Map project. OSM uses the WGS84 CRS, so we will assume that the example datasets do too.

```{r}
metadata <- xml2::read_xml(here("map-matching-dataset/metadata.xml"))

metadata_xml <- XML::xmlParse(metadata)
```

```{r}
# Track 0 is somewhere in Russia
track0 <- read_tsv(here("map-matching-dataset/00000000/00000000.track"), col_names = FALSE)

track0 <- track0 |>
  rename(lon = "X1",
         lat = "X2",
         time = "X3")

track0_sf <- track0 |>
  st_as_sf(coords = c("lon", "lat"), 
           crs = "WGS84")
```

```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data = track0_sf, label = ~glue("{time}"))
```

```{r}
# Road segment endpoints
nodes0 <- read_tsv(here("map-matching-dataset/00000000/00000000.nodes"), col_names = FALSE)

nodes0 <- nodes0 |>
  rename(lon = "X1",
         lat = "X2") |>
  mutate(node_id = seq(0, nrow(nodes0) - 1, 1))

nodes0_sf <- nodes0 |>
  st_as_sf(coords = c("lon", "lat"), 
           crs = "WGS84")
```

```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data = track0_sf, label = ~glue("{time}"), color = "blue") |>
  addCircleMarkers(data = nodes0_sf, color = "red")
```

```{r}
# Road segment connections
arcs0 <- read_tsv(here("map-matching-dataset/00000000/00000000.arcs"), col_names = FALSE)

arcs0 <- arcs0 |>
  rename(start_node = "X1",
         end_node = "X2")
```

```{r}
# Sequence of arcs on which the vehicle traveled
# Each row refers to an arc in the arcs table using its (zero-based) row number.
route0 <- read_tsv(here("map-matching-dataset/00000000/00000000.route"), col_names = FALSE)
```

```{r}
test_start <- nodes0_sf |> filter(node_id == 0)
test_end <- nodes0_sf |> filter(node_id == 2433)

test_ls <- st_cast(st_union(test_start, test_end, by_feature = TRUE), "LINESTRING")
```

```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data = test_start, color = "red", label = "0") |>
  addCircleMarkers(data = test_end, color = "purple", label = "2433") |>
  addPolylines(data = test_ls, color = "black")
```

```{r}
test_start
lwgeom::st_startpoint(test_ls)
# These don't match
```

```{r}
test_end
lwgeom::st_endpoint(test_ls)
# These don't match
```

```{r}
# Try a different way to make lwgeom::st_startpoint and lwgeom::st_endpoint behave as expected
test_ls2 <- st_linestring(rbind(st_coordinates(test_start), st_coordinates(test_end))) |>
  st_sfc(crs = "WGS84")

test_ls2_df <- data.frame(start_node = 0, end_node = 2433) |>
  st_set_geometry(test_ls2)
```

```{r}
test_start
lwgeom::st_startpoint(test_ls2)
# These match
```

```{r}
test_end
lwgeom::st_endpoint(test_ls2)
# These match
```


```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data = test_start, color = "red", label = "0") |>
  addCircleMarkers(data = test_end, color = "purple", label = "2433") |>
  addPolylines(data = test_ls, color = "gray")
```

```{r}
road_segments0 <- map2(
  .x = arcs0$start_node, .y = arcs0$end_node,
  \(x, y) data.frame(start_node = x, end_node = y) |>
    st_set_geometry(st_linestring(rbind(st_coordinates(nodes0_sf |> filter(node_id == x)), st_coordinates(nodes0_sf |> filter(node_id == y)))) |>
  st_sfc(crs = "WGS84"))
)

road_segments0_df <- bind_rows(road_segments0)

road_segments0_df <- road_segments0_df |>
  mutate(arc_id = seq(0, nrow(road_segments0_df) - 1, 1))
```

```{r}
all.equal(arcs0$start_node, road_segments0_df$start_node) # TRUE

all.equal(arcs0$end_node, road_segments0_df$end_node) # TRUE
```

```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data = track0_sf, label = ~glue("{time}"), color = "blue") |>
  addCircleMarkers(data = nodes0_sf, color = "red") |>
  addPolylines(data = road_segments0_df, color = "gray", label = ~glue("Start: {start_node}, End: {end_node}"))
```

```{r}
route0 <- route0 |>
  rename(arc_id = "X1")
```

```{r}
route_arcs <- left_join(route0, road_segments0_df, by = "arc_id") |> st_as_sf()
```

```{r}
leaflet() |>
  addTiles() |>
  addCircleMarkers(data = track0_sf, label = ~glue("{time}"), color = "blue") |>
  #addCircleMarkers(data = nodes0_sf, color = "red") |>
  addPolylines(data = route_arcs, color = "gray", label = ~glue("Start: {start_node}, End: {end_node}"))
```
Find a track in the United States
```{r}
track <- read_tsv(here("map-matching-dataset/00000090/00000090.track"), col_names = FALSE)

track <- track |>
  rename(lon = "X1",
         lat = "X2",
         time = "X3")

track_sf <- track |>
  st_as_sf(coords = c("lon", "lat"), 
           crs = "WGS84")
```

```{r}
leaflet() |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircleMarkers(data = track_sf, label = ~glue("{time}"))
```
* Track 1 is in Russia.

* Track 2 is in Germany.

* Track 3 is in Germany.

* Track 4 is in Russia.

* Track 5 is in Ukraine.

* Track 6 is in Ukraine.

* Track 7 is in Kazakhstan.

* Track 8 is in Alabama.

* Track 9 is in Italy.

* Track 10 is in Russia.

* Track 20 is in Belarus.

* Track 30 is in Spain.

* Track 40 is in Russia.

* Track 50 is in Germany.

* Track 60 is in Poland.

* Track 70 is in France.

* Track 80 is in Argentina.

* Track 90 is in Germany.
